Parameters associated with Injuries in the NFL

### 2024-06-22 products/manuscript/Manuscript.qmd
Submission part 3 changes

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

@@

Three data sets have been identified for this project. The primary data set ("Injury Player Data" obtained from a Github repository contain 1,586 observations) consists of every play ran in the NFL containing an injury during the 2019-2020 season. Some of the things contained in this data include the teams playing, the week in the season, play description, injury area, and injured player metrics. Two other data sets were obtained from Kaggle. The first is a "cumulative play data" containing information about every play ran in the 2019-2020 NFL season (42,186 observations). Information such as the teams playing, play description, and time in the game when it was ran. The last data set is the "player demographic data" containing information about all NFL players in the 2015-2019 seasons (11,145 observations). This data set has player metrics such as height, weight, date of birth, etc.

@@

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

@@

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

In both the "Injury Player Data" set and "cumulative play data" set, the play descriptions will need to be boiled down to simplified play types (pass, run, fumble, etc.). Next, a lot of what looks like unnecessary plays are included in the "cumulative play data" set such as end of quarters and games so these observations can be filtered out. Lastly, empty columns were noticed in the "cumulative play data" that can be removed.

@@

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

```{r}
#Loading the Data Sets
#install.packages("RCurl")
#library(RCurl)

# Injury Player Data
#injuries <- read.csv(text = getURL("https://raw.githubusercontent.com/btj5z2/BrodyJohnson-P2-portfolio/main/position_injury_player_2019_2020_update.csv")) #also works but not needed
injuries=read.csv("../../data/raw-data/position_injury_player_2019_2020_update.csv")

# Cumulative Play data
#pbp <- read.csv(text = getURL("https://raw.githubusercontent.com/btj5z2/BrodyJohnson-P2-portfolio/main/pbp-2019.csv")) #short for play-by-play #also works but not needed
pbp = read.csv("../../data/raw-data/pbp-2019.csv")

#Player Demographic data
#players <- read.csv(text = getURL("https://raw.githubusercontent.com/btj5z2/BrodyJohnson-P2-portfolio/main/player_position_Rmd1.csv"))
players = read.csv("../../data/raw-data/players (1).csv")

## Statistical analysis
_Explain anything related to your statistical analyses._
The big picture look at the analysis will be to compare the frequency of parameters in the injury data set to the same parameters in the two supporting data sets. For example, we may find in the injury data set that most injuries occur during "short pass" plays. However, if we were to find the "short pass" play is the most common play type ran in the NFL, this becomes less significant of a finding.

@@

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@@

```{r}
#Structure of Injury Data Set
str(injuries)
```

```{r}
#Structure of Play-by-Play Data
str(pbp)
```

```{r}
#Structure of All Players Data Set
str(players)
```

@@

```{r}
hist(injuries$week.x, breaks=24,main="Injuries by Week in Regular Season", xlim=c(1,18),xlab="Week",labels =T)
```

@@

```{r}
par(mfrow=c(1,2))
hist(injuries$qtr, breaks=6,main="Injuries by Quarter", xlim=c(0,5),xlab="Quarter",labels =T)
hist(pbp$Quarter, breaks=6,main="Total Plays by Quarter", xlim=c(0,5), xlab="Quarter",labels =T)

@@

```{r}
par(mfrow=c(1,2))
hist(injuries$down, breaks=6,main="Injuries by Down", xlim=c(0,5),xlab="Quarter",labels =T)
hist(pbp$Down, breaks=6,main="Total Plays by Down", xlim=c(0,5), xlab="Quarter",labels =T)

@@

```{r}
par(mfrow=c(1,2))
hist(injuries$height, breaks=10,main="Injuries by Height", xlab="Heiht",labels =T)
hist(as.numeric(players$height), breaks=10,main="Total Players Height", xlab="Height",labels =T) #Nedd to look into why not all numeric

@@

```{r}
par(mfrow=c(1,2))
hist(injuries$weight, breaks=10,main="Injuries by Weight", xlab="Weight",labels =T)
hist(players$weight, breaks=10,main="Total Players Weight", xlab="Weight",labels =T)

@@

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

@@

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

>>>>

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(dplyr)
library(tidyverse)
```

@@

Three data sets have been identified for this project. The primary data set ("Injury Player Data" obtained from a Github repository contains 1,586 observations) consists of every play ran in the NFL containing an injury during the 2019-2020 season. Some of the things contained in this data include the teams playing, the week in the season, play description, injury area, and injured player metrics. Two other data sets were obtained from Kaggle. The first is a "cumulative play data" containing information about every play ran in the 2019-2020 NFL season (42,186 observations). Information such as the teams playing, play description, and time in the game when it was ran. The last data set is the "player demographic data" containing information about all NFL players in the 2016-2019 seasons (11,145 observations). This data set has player metrics such as height, weight, date of birth, etc.

@@

*****To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].******

@@

The primary idea behind the analysis for this project is to look at various parameters when an injury occurred and compare them to the rest of the NFL. If the parameters are related to the play itself, the paramter in the injruy data set will be compared to cumulative play data set. Likewise, if the paramter is related to the injured player themselves, this paramter will be copmared to the cumulative player data set. If there is a statistical difference between paramters, we can likely draw some conclusion about a parameter causing injuries.

With this being said, it is imperative to ensure parameters from both data sets are in the same format in order to compare them. Other data cleaning items will include vetting the data sets for unexpected/missing values and converting variables to another variable that easier to use. For example, the play descriptions in the injury data set and cumulative play data set are very extensive in what happened during the play. Therefore, this will be boiled down to be either a run play, short pass, etc.

@@

## Data import and cleaning
In both the "Injury Player Data" set and "cumulative play data" set, the play descriptions will need to be boiled down to simplified play types (pass, run, fumble, etc.). Next for the "Injury Player Data" set, the "injury_area" parameter will be summarized as upper and lower body injuries. This is due to the overwhelming number of injuries being classified as one or the other already so this will classify the remaining injuries as such. For the injury data set and the "player demographic data", player ages will need to be calculated from their birth-dates.

As for the "cumulative player data" set, empty columns will be removed and 0's in the down column will be converted to "NA" values which indicate kickoffs & PATs. Lastly, a lot of unnecessary plays are included in the "cumulative play data" set such as end of quarter/games, 2-minute warnings, and timeouts so these observations can be filtered out.

In the "player demographic data" set, observations for players height were recorded in both inches and feet-inches so this was cleaned to only contain data in inches. Lastly, the players position was converted to a factor for analysis.

```{r}
#Loading the processed Data Sets
#Raw data files were cleaned in the processingcode.R (../../code/processing-code/) then saved as .rds file in the processed-data folder

# Injury Player Data
injuries=readRDS("../../data/processed-data/injuriesprocesseddata.rds")

# Cumulative Play data
pbp=readRDS("../../data/processed-data/pbpprocesseddata.rds")

#Player Demographic data
players=readRDS("../../data/processed-data/playersprocesseddata.rds") 
```

## Statistical analysis
First, I will select variables that show a potential difference between the injury data set and cumulative  data sets through exploratory analysis. From these variables, I will run  inferential tests for the numeric variables and a Chi-square test for the categorical variables.

@@

@@

```{r}
#Structure of Injury Data Set
glimpse(injuries)
```

```{r}
#Structure of Play-by-Play Data
glimpse(pbp)
```

```{r}
#Structure of All Players Data Set
glimpse(players)
```

@@

Looking at the number of injuries each week in the regular season, there's a sharp drop after the first week, steadily climbs nearing the end of the season, and another drop at the end of the season. Intuitively, this makes sense in that players may not be physically readily for the intensity of a regular-season game despite training camps and preseason games. After that, the slight rise in the second half of the season could be due to exhaustion through a long season and/or teams fighting more for playoff spots. Lastly, the steep drop in the last week is likely due to teams knowing whether or not they have made playoffs and therefore, do not want to incur an injury to go into either the post-season or off-season.  
```{r}
hist(injuries$week.x, breaks=21,main="Injuries by Week in Regular Season", xlim=c(1,17),ylim=c(0,200),xlab="Week",labels =T)
```

Below depicts the number of injuries for each reported injury area. The top results were simply reported as upper or lower body.

```{r}
injarea_freq=injuries %>%  group_by(injury.area) %>% summarize(count=n()) 
ggplot(data = injarea_freq, aes(x = reorder(injury.area, -count), y = count))+
  geom_col() +
  labs(x = "Injury Area", y="# of Injuries") + 
  theme(axis.text.x = element_text(angle=45, hjust = 1))

Below is a plot of the number of injuries by injury area but the injury areas were summarized to either be upper/lower body or head injuries. I kept head injuries separate due to the interest around concussions in football and see how they stack up against all other injuries.

```{r}
injareanew_freq=injuries %>%  group_by(injury.area.new) %>% summarize(count=n()) 
ggplot(data = injareanew_freq, aes(x = reorder(injury.area.new, -count), y = count))+
  geom_col() +
  labs(x = "Injury Area", y="# of Injuries") + 
  theme(axis.text.x = element_text(angle=45, hjust = 1))
```

Out of curiosity, I plotted below the frequency of injuries for each team during the regular season. I was not expecting much but it is interesting to see about a triple in number of injuries from the lowest team (LA Rams) and the highest (LA Chargers).

```{r}
team_freq=injuries %>%  filter(week.x<18) %>% group_by(team) %>% summarize(count=n()) #filtered to only contain regular season games (weeks 1-17)
ggplot(data = team_freq, aes(x = reorder(team, -count), y = count))+
  geom_col() +
  labs(x = "Team", y="# of Injuries") + 
  theme(axis.text.x = element_text(angle=45, hjust = 1))
```

@@

The number of plays seem evenly distributed among the halves (slight increase in 2nd and 4th quarters likely due to running last minute plays). However, the number of injuries consistently increase throughout the game which also makes sense as players fatigue.

```{r}
hist(injuries$qtr,main="Injuries by Quarter", ylim=c(0,550), xlim=c(1,5),xlab="Quarter",labels =T)

```
```{r}
hist(pbp$Quarter, breaks=6,main="Total Plays by Quarter", ylim=c(0,11200), xlab="Quarter",labels =T)

@@

There appears to be a very similar trend among the number of plays per down and the number of injuries ocurring during each down. That is, they both decrease at similar rates.

```{r}
hist(injuries$down, breaks=5,main="Injuries by Down", ylim=c(0,700),xlim=c(1,4),xlab="Down",labels =T)
```

```{r}
hist(pbp$Down, breaks=6,main="Total Plays by Down", ylim=c(0,14900),xlim=c(1,4), xlab="Down",labels =T)
```

Below are two histograms depicting first the number of injuries and then second the total # of plays by different play types. After a glance, there does not appear to be a significant difference in number of injuries and total # of plays ran by play type.

```{r}
play_type_freq=injuries %>%  group_by(play_type) %>% summarize(count=n()) 
ggplot(data = play_type_freq, aes(x = reorder(play_type, -count), y = count))+
  geom_col() +
  labs(x = "Play Type", y="# of Injuries") + 
  theme(axis.text.x = element_text(angle=45, hjust = 1))

```{r}
play_type_freq2=pbp %>%  group_by(play_type) %>% summarize(count=n()) 
ggplot(data = play_type_freq2, aes(x = reorder(play_type, -count), y = count))+
  geom_col() +
  labs(x = "Play Type", y="Total # of Plays") + 
  theme(axis.text.x = element_text(angle=45, hjust = 1))

@@

Below are the distributions of players height first of the injured players and then from the cumulative player data set. Both histograms appear to be normally distributed. One interesting difference is the injured players appear to have more players just below and above the center of the histogram. That is, it would almost suggest players specifically under and above the mean height have a higher inclination for injury.

```{r}
hist(injuries$height, breaks=10,main="Injuries by Height", ylim=c(0,250), xlab="Height",labels =T)

```
```{r}
hist(players$height, breaks=10,main="Total Players Height", ylim=c(0,250), xlab="Height",labels =T)

@@

Below are the histograms depicting the players height. They are both skewed to the right with one interesting finding of a peak of injured players in the ~300lb range.

```{r}
hist(injuries$weight, breaks=10,main="Injuries by Weight", xlim=c(150,400),ylim=c(0,400), xlab="Weight",labels =T)
```

```{r}
hist(players$weight, breaks=10,main="Total Players Weight", xlim=c(150,400),ylim=c(0,400),  xlab="Weight",labels =T) 
```

Both distributions of players age are below and have a similar skewness to the right.

```{r}
hist(injuries$age, breaks=10,main="Injuries by Age", xlim=c(20,45),ylim=c(0,425), xlab="Age",labels =T)
```

```{r}
hist(players$age, breaks=10,main="Total Players Age", xlim=c(20,45),ylim=c(0,425),  xlab="Age",labels =T) 
```

The primary finding from the player positions is that it appears the linemen (defensive and ofensive) are injured more often relative to how many there are in the league.

```{r}
position_freq=injuries %>%  group_by(position) %>% summarize(count=n()) 
ggplot(data = position_freq, aes(x = reorder(position, -count), y = count))+
  geom_col() +
  labs(x = "Player Position", y="# of Injuries") + 
  theme(axis.text.x = element_text(angle=45, hjust = 1))

@@

```{r}
position_freq2=players %>%  group_by(position) %>% summarize(count=n()) 
ggplot(data = position_freq2, aes(x = reorder(position, -count), y = count))+
  geom_col() +
  labs(x = "Player Position", y="Total # of Players") + 
  theme(axis.text.x = element_text(angle=45, hjust = 1))

@@

From the exploratory analysis, it appears there three numeric variables that may show a difference and one categorical. The three numeric variables are quarter, height, and weight while the single categorical variable is player position.

```{r}
#Inferential test looking at Quarter
shapiro.test(injuries$qtr) #p-value<0.05. Reject Ho & don't assume normal.
shapiro.test(sample(pbp$Quarter, size=5000, replace=F)) #p-value<0.05. Reject Ho & don't assume normal.
#Will do Wilcoxon rank-sum test b/c of non-normal data

wilcox.test(injuries$qtr, pbp$Quarter, exact=F) #p-value<0.05. Reject Ho and therefore, one group tends to have larger values than the other

```{r}
#Inferential test looking at Height
shapiro.test(injuries$height) #p-value<0.05. Reject Ho & don't assume normal.
shapiro.test(players$height) #p-value<0.05. Reject Ho & don't assume normal.
#Will do Wilcoxon rank-sum test b/c of non-normal data

wilcox.test(injuries$height, players$height, exact=F) #p-value<0.05. Reject Ho and therefore, one group tends to have larger values than the other

```{r}
#Inferential test looking at Weight
shapiro.test(injuries$weight) #p-value<0.05. Reject Ho & don't assume normal.
shapiro.test(players$weight) #p-value<0.05. Reject Ho & don't assume normal.
#Will do Wilcoxon rank-sum test b/c of non-normal data

wilcox.test(injuries$weight, players$weight, exact=F) #p-value<0.05. Reject Ho and therefore, one group tends to have larger values than the other

```
```{r}
#Chi-square test looking at player position
factor_levels = levels(injuries$position)
chisq.test(sample(injuries$position, size=1303, replace=F), players$position) #p-value>0.05. Accept Ho and therefore, the two data sets have equal proportions among the player positions
